---
title: "proj_1"
output:
  html_document:
    self_contained: no
    keep_md: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
```

## Load the data 


```{r readData}
zipfile_url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
download.file(zipfile_url, destfile = ".\\zip_file.zip")
unzip('zip_file.zip')
file_name <- 'activity.csv'
data <- read.csv(file_name)
```


## Process the data into a format suitable for your analysis
Create new column with date as a Date object. 
```{r makeDate}
data$date_d <- as.Date(as.character(data$date))
```


## mean total number of steps taken per day

```{r stepsByDay}
steps_Day <- with(data, tapply(steps, date, sum, na.rm = T)) 
```
Check that there are 61 rows (days).
```{r}
nrow(steps_Day) 
```

## Make a histogram of the total number of steps taken each day.
```{r histogram, fig.width = 10, fig.height=6, fig.path='Figures/'}
hist(steps_Day, breaks=seq(0,22500, 2500), xaxt = 'n', xlab = "Total Steps Taken", 
     ylab = "Number of Days", ylim = range(seq(0,20, 5)), main = '', col = 'powderblue')
axis(side=1, at=seq(0,22500, 2500), labels=seq(0,22500, 2500))
```

## Calculate and report the mean and median of the total number of steps taken per day
```{r MeanMedian}
mean_steps <- round(mean(steps_Day), digits = 3)
median_steps <- round(median(steps_Day), digits = 3)
mean_steps
median_steps
```

## calculate the average daily activity pattern  

```{r stepsByInt}
steps_byInt <- with(data, tapply(steps, interval, mean, na.rm = T)) 
```

Check that there are 288 rows (1440 minutes in a day, split into 5 minute intervals).
```{r}
nrow(steps_byInt) 
```

Create plot.
```{r linePlot, fig.path='Figures/'}
plot(steps_byInt, type = 'l', xlab = 'Interval Number', ylab = 'Average Number of Steps Taken')
```

## Which 5-minute interval, on average across all the days in the dataset?
```{r maxInt}
max_int <- match(max(steps_byInt), steps_byInt)
max_int
```
The interval with the maximum number of steps, on average, is interval `r max_int`, which corresponds to minute `r data$interval[104]`.

## Calculate and report the total number of missing values in the dataset 
```{r missingValues}
missing <- sum(is.na(data$steps))
missing
```
The total number of missing values is `r missing`.

```{r}
data_filled <- data
for(i in 1:nrow(data_filled)){
    if(is.na(data_filled$steps[i])){
        int <- as.character(data_filled$interval[i])
        num <- steps_byInt[[int]]
        data_filled$steps[i] <- num
    }
}
```


## Make the histogram 
```{r}
steps_byDay_2 <- with(data_filled, tapply(steps, date, sum)) 
```

Check that there are 61 rows (days).
```{r}
nrow(steps_byDay_2) 
```
```{r histogram2, fig.width=10, fig.height=6, fig.path='Figures/'}
hist(steps_byDay_2, breaks=seq(0,22500, 2500), xaxt = 'n', xlab = "Total Steps Taken", 
     ylab = "Number of Days", ylim = range(seq(0, 30, 5)), main = '', col = 'powderblue')
axis(side=1, at=seq(0,22500, 2500), labels=seq(0,22500, 2500))
```

## Calculate and report the mean and median total number of steps taken per day
```{r MeanMedian2}
mean_steps2 <- round(mean(steps_byDay_2), digits = 3)
median_steps2 <- round(median(steps_byDay_2), digits = 3)
mean_steps2
median_steps2
```
The mean total number of steps per day is `r mean_steps2`. The median total number of steps per day is `r median_steps2`.


**Are there differences in activity patterns between weekdays and weekends?**  
**For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.**  

**1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.**

```{r createFactor}
weekend <- c('Saturday','Sunday')
for(i in 1:nrow(data_filled)){
    day <- weekdays(data_filled$date_d[i])
    if(day %in% weekend){
        data_filled$day_type[i] <- 'weekend'
    } else {
        data_filled$day_type[i] <- 'weekday'
    }
}
data_filled$day_type <- as.factor(data_filled$day_type) 
```
Check new variable and data.
```{r checkFactor}
class(data_filled$day_type)
str(data_filled)
```

**2. Make a panel plot containing a time series plot (i.e. type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).**  


```{r subsetData}
weekday_data <- subset(data_filled, data_filled$day_type == 'weekday')
weekday_steps_data <- with(weekday_data, tapply(steps, c(interval), mean, na.rm = T)) 

weekend_data <- subset(data_filled, data_filled$day_type == 'weekend')
weekend_steps_data <- with(weekend_data, tapply(steps, c(interval), mean, na.rm = T)) 
```

Check data; nrows should be 288.
```{r checkData}
head(weekday_steps_data)
head(weekend_steps_data)
nrow(weekday_steps_data)
nrow(weekend_steps_data)
```

Create plot.
```{r panelPlot, fig.path='Figures/'}
par(mfrow=c(1,2))
plot(weekday_steps_data, type = 'l', xlab = 'Interval Number', ylab = 'Average Number of Steps Taken', 
     main = 'Weekdays', ylim = c(0,250))
plot(weekend_steps_data, type = 'l', xlab = 'Interval Number', ylab = 'Average Number of Steps Taken', 
     main = 'Weekends',ylim = c(0,250))
```
